---
output:
  pdf_document: default
  html_document: default
---
# Lesson 1: Built-in Functions

* Understanding the purpose and usage of common built-in R functions (e.g., `mean()`, `sum()`, `max()`, `min()`, `length()`).
* Function arguments and default values.
* Using R's help system (`?`, `help()`, `example()`).

R comes with a vast collection of built-in functions that allow you to perform a wide variety of tasks without needing to write complex code yourself. These functions are the workhorses of R, enabling everything from simple calculations to complex statistical analyses.

---

## **Phase 1: Introduction to R and Fundamentals**

### **Module 1.4: Functions and Packages**

---

### **Lesson 1: Built-in Functions**

This lesson introduces you to some of the most commonly used and important built-in functions in R. These functions provide powerful capabilities for performing mathematical operations, statistical analysis, string manipulation, and data structure manipulation.

---

#### **1. Common Mathematical Functions**

These functions are used for basic mathematical calculations on numeric data.

* `sum()`: Calculates the sum of all elements in a numeric vector.
* `mean()`: Calculates the arithmetic mean (average) of a numeric vector.
* `median()`: Calculates the median of a numeric vector.
* `min()`: Finds the minimum value in a numeric vector.
* `max()`: Finds the maximum value in a numeric vector.
* `sqrt()`: Calculates the square root of numbers.
* `abs()`: Calculates the absolute value of numbers.
* `round()`: Rounds numbers to a specified number of decimal places.

**Code Snippets:**

```{r}
# Create a numeric vector for examples
numbers <- c(10, 25, 5, 30, 15, NA, 20)

print("--- Common Mathematical Functions ---")
```
#### sum()
```{r}

print(paste("1. Sum of numbers:", sum(numbers, na.rm = TRUE))) # na.rm=TRUE removes NA for calculation
print(paste("   Sum of numbers (without na.rm):", sum(numbers)))
```
#### mean()
```{r}

print(paste("2. Mean of numbers:", mean(numbers, na.rm = TRUE)))
print(paste("   Mean of numbers (without na.rm):", mean(numbers)))
```
#### median()
```{r}
print(paste("3. Median of numbers:", median(numbers, na.rm = TRUE)))
```
#### min()
```{r}

print(paste("4. Minimum number:", min(numbers, na.rm = TRUE)))
```
#### max()
```{r}

print(paste("5. Maximum number:", max(numbers, na.rm = TRUE)))
```
#### sqrt()
```{r}
positive_num <- 16
print(paste("6. Square root of", positive_num, ":", sqrt(positive_num)))
vec_for_sqrt <- c(4, 9, 25)
print(paste("   Square roots of c(4,9,25):", paste(sqrt(vec_for_sqrt), collapse = ", ")))
```
#### abs()
```{r}
negative_num <- -7.5
print(paste("7. Absolute value of", negative_num, ":", abs(negative_num)))
vec_for_abs <- c(-10, 5, -2)
print(paste("   Absolute values of c(-10,5,-2):", paste(abs(vec_for_abs), collapse = ", ")))
```
#### round()
```{r}

decimal_num <- 12.3456
print(paste("8. Round 12.3456 to 2 decimal places:", round(decimal_num, digits = 2)))
print(paste("   Round 12.3456 to nearest integer:", round(decimal_num)))
```

---

#### **2. Statistical Functions**

R provides a rich set of functions for basic statistical analysis.

* `sd()`: Calculates the standard deviation.
* `var()`: Calculates the variance.
* `summary()`: Provides a statistical summary of a vector or data frame.
* `quantile()`: Calculates quantiles (e.g., 25th, 50th, 75th percentiles).

**Code Snippets:**

```{r}
# Create a numeric vector for examples
data_points <- c(10, 12, 15, 13, 18, 11, 14)
scores_with_na <- c(85, 92, 78, NA, 65, 95, 88)

print("--- Statistical Functions ---")
```
#### sd() - Standard Deviation
```{r}

print(paste("1. Standard deviation of data_points:", round(sd(data_points), 2)))
print(paste("   Standard deviation of scores_with_na (na.rm=TRUE):", round(sd(scores_with_na, na.rm = TRUE), 2)))
```
#### var() - Variance
```{r}
print(paste("2. Variance of data_points:", round(var(data_points), 2)))
```
#### summary() - Provides a statistical summary
```{r}
print("3. Summary of data_points:")
summary(data_points)
print("   Summary of a data frame:")
my_df <- data.frame(
    ID = 1:3,
    Name = c("A", "B", "C"),
    Value = c(10, 20, 30),
    stringsAsFactors = FALSE
)
summary(my_df)
```

# quantile()
```{r}
print("4. Quantiles of data_points:")
print(quantile(data_points)) # Default quantiles (0%, 25%, 50%, 75%, 100%)
print("   Specific quantiles (10%, 90%):")
print(quantile(data_points, probs = c(0.10, 0.90)))
```

#### **3. Character Manipulation Functions**

These functions are used for working with text (character strings).

* `nchar()`: Counts the number of characters in a string or each element of a character vector.
* `paste()`: Concatenates (joins) strings. Can be used with `paste0()` for no separator.
* `grep()`: Searches for matches to a regular expression (pattern) within character vectors.
* `sub()`: Replaces the *first* occurrence of a pattern in strings.
* `gsub()`: Replaces *all* occurrences of a pattern in strings.

**Code Snippets:**

```{r}
# Create character vectors for examples
text1 <- "Hello R"
names_vec <- c("Alice", "Bob", "Charlie", "David")
sentence <- "R is a powerful statistical language, R is great."

print("--- Character Manipulation Functions ---")
```
# nchar()
```{r}
print(paste("1. Number of characters in '", text1, "':", nchar(text1)))
print(paste("   Characters in names_vec:", paste(nchar(names_vec), collapse = ", ")))
```
#### paste() / paste0()
```{r}
combined_text_space <- paste("Welcome", "to", "R")
print(paste("2. Using paste() (default sep=' '):", combined_text_space))
combined_text_no_space <- paste0("Welcome", "to", "R")
print(paste("   Using paste0() (no separator):", combined_text_no_space))
combined_with_sep <- paste("Item", 1:3, sep = "-")
print(paste("   Paste with separator (sep='-'):", paste(combined_with_sep, collapse = ", ")))
```
```{r}
#### grep() - find elements matching a pattern
# Returns indices by default, use value=TRUE to get the values
search_results_idx <- grep("a", names_vec)
print(paste("3. Indices of names containing 'a':", paste(search_results_idx, collapse = ", ")))
search_results_val <- grep("a", names_vec, value = TRUE)
print(paste("Names containing 'a':", paste(search_results_val, collapse = ", ")))
```
#### sub() - replace first occurrence
```{r}

replaced_first <- sub("R", "Python", sentence)
print(paste("4. Replace first 'R' with 'Python':", replaced_first))
```
#### gsub() - replace all occurrences
```{r}
replaced_all <- gsub("R", "Python", sentence)
print(paste("5. Replace all 'R' with 'Python':", replaced_all))
```

---

#### **4. Vector/Matrix Manipulation Functions**

These functions are used to get information about or manipulate vectors and matrices.

* `length()`: Returns the number of elements in a vector.
* `dim()`: Returns the dimensions (rows, columns) of a matrix or data frame.
* `nrow()`: Returns the number of rows of a matrix or data frame.
* `ncol()`: Returns the number of columns of a matrix or data frame.
* `t()`: Transposes a matrix or data frame.
* `c()`: Combines vectors or lists into a new vector.
* `rbind()`: Combines vectors, matrices, or data frames by rows.
* `cbind()`: Combines vectors, matrices, or data frames by columns.

**Code Snippets:**
#### Create vectors and matrices for examples

```{r}
my_vec <- c(10, 20, 30, 40, 50)
my_matrix <- matrix(1:9, nrow = 3, byrow = TRUE)
vec_to_combine_1 <- c("A", "B")
vec_to_combine_2 <- c("C", "D")

print("--- Vector/Matrix Manipulation Functions ---")
```

# length()
```{r}
print(paste("1. Length of my_vec:", length(my_vec)))
```
#### dim()
```{r}

print("2. Dimensions of my_matrix (rows, cols):")
print(dim(my_matrix))
```
#### nrow()
```{r}

print(paste("3. Number of rows in my_matrix:", nrow(my_matrix)))
```
#### ncol()
```{r}

print(paste("4. Number of columns in my_matrix:", ncol(my_matrix)))
```
#### t() - Transpose
```{r}
print("5. Original Matrix:")
print(my_matrix)
print("   Transposed Matrix:")
print(t(my_matrix))
```
#### c() - Combine elements (coerces to common type)
```{r}
combined_vec <- c(my_vec, 60, 70)
print(paste("6. Combined vector (c()):", paste(combined_vec, collapse = ", ")))
mixed_type_c <- c(1, "hello", TRUE) # Coerces all to character
print(paste("   Mixed type combined with c():", paste(mixed_type_c, collapse = ", ")))
print(paste("   Class of mixed_type_c:", class(mixed_type_c)))

```
# rbind() - Row bind (combines by rows)
```{r}
row1 <- c(1, 2, 3)
row2 <- c(4, 5, 6)
combined_rows <- rbind(row1, row2)
print("7. Row-bound matrix:")
print(combined_rows)
```
# cbind() - Column bind (combines by columns)
```{r}
col1 <- c(10, 20)
col2 <- c(30, 40)
combined_cols <- cbind(col1, col2)
print("8. Column-bound matrix:")
print(combined_cols)

# Note: rbind/cbind can also be used with data frames, 
# but require same number of columns/rows respectively.
```

#### **5. Data Frame Specific Functions**

These functions are particularly useful for inspecting and working with data frames, the most common data structure for tabular data.

* `head()`: Displays the first few rows of a data frame (default 6).
* `tail()`: Displays the last few rows of a data frame (default 6).
* `str()`: Displays the internal structure of an R object, very useful for data frames to see column names and types.
* `summary()`: Provides a statistical summary of each column in a data frame.
* `colnames()`: Gets or sets the column names of a data frame.
* `rownames()`: Gets or sets the row names of a data frame.

**Code Snippets:**
#### Create a data frame for examples
```{r}
employee_data <- data.frame(
    EmpID = c("E001", "E002", "E003", "E004", "E005"),
    Name = c("John", "Jane", "Peter", "Mary", "Chris"),
    Department = c("HR", "IT", "Sales", "HR", "IT"),
    Salary = c(60000, 85000, 70000, 62000, 90000),
    YearsExp = c(5, 10, 7, 6, 12),
    stringsAsFactors = FALSE
)

print("--- Data Frame Specific Functions ---")
print("Original Data Frame:")
print(employee_data)
```
# head()
```{r}
print("1. First 3 rows of employee_data:")
print(head(employee_data, n = 3))
```
#### tail()
```{r}
print("2. Last 2 rows of employee_data:")
print(tail(employee_data, n = 2))
```
#### str()
```{r}
print("3. Structure of employee_data (str()):")
str(employee_data)
```

# summary()
```{r}
print("4. Summary of employee_data:")
summary(employee_data)
```

# colnames() - Get column names
```{r}
print("5. Column names of employee_data:")
print(colnames(employee_data))
# colnames() - Set column names (example: rename 'YearsExp')
colnames(employee_data)[5] <- "YearsExperience"
print("   Updated column names:")
print(colnames(employee_data))
print("   Data frame with updated column name:")
print(employee_data) # Show effect of renaming
```
# rownames() - Get row names (usually just numbers by default)
```{r}
print("6. Row names of employee_data:")
print(rownames(employee_data))
# rownames() - Set row names (e.g., to EmpID)
# rownames(employee_data) <- employee_data$EmpID
# print("   Updated row names (using EmpID):")
# print(rownames(employee_data))
```

This lesson provided a foundational understanding of many common and essential built-in functions in R, covering mathematical, statistical, character, vector/matrix, and data frame operations. Familiarity with these functions will significantly enhance your ability to perform data manipulation and analysis in R.

**Next, we will proceed to Lesson 2: Creating Custom Functions.**
