# Lesson 3: Lists

* Understanding lists as heterogeneous data structures.
* Creating lists (`list()`).
* Accessing list elements using `[]`, `[[]]`, and `$`.
* Nested lists.

**Module 1.2: Data Types and Structures**.

Lists are incredibly versatile in R, as they allow you to store collections of objects of *different* types, unlike atomic vectors and matrices/arrays. This makes them a powerful tool for organizing various pieces of information.

---

## **Phase 1: Introduction to R and Fundamentals**

### **Module 1.2: Data Types and Structures**

---

### **Lesson 3: Lists**

This lesson introduces **lists**, a highly flexible data structure in R. Unlike atomic vectors, matrices, or arrays, lists can hold objects of *different* data types and even different structures (e.g., a list can contain a numeric vector, a character string, a logical value, and even another list or a data frame). This makes them incredibly useful for organizing complex data.

---

#### **Understanding Lists: Heterogeneous Data Containers**

A **list** in R is an ordered collection of objects. The key characteristic of a list is its **heterogeneity**: each element (or component) of a list can be a different type of R object – it could be an atomic vector, a matrix, a data frame, a function, or even another list.

Think of a list as a generic container that can hold various items, each potentially unique in its form and content.

**Key Characteristics of Lists:**

* **Heterogeneous:** Can store elements of different data types (numeric, character, logical, etc.) and different R object types (vectors, matrices, data frames, other lists, functions, etc.).
* **Ordered:** Elements are stored in a specific order and can be accessed by their position.
* **Named or Unnamed:** Elements can be named, which greatly improves readability and access.

---

#### **Creating Lists**

The primary function to create a list in R is `list()`.

**Syntax:**

```{r}
list(element1, element2, ..., elementN)
```

**Parameters:**

* `element1, element2, ...`: The objects you want to include as components of the list. These can be any R objects.

**Return Value:**

* A list object.

**Code Snippets:**

```{r}
# Create a simple list with unnamed elements
my_first_list <- list("apple", 10, TRUE)
print("My First List (unnamed):")
print(my_first_list)

# Create a list with named elements (recommended for clarity)
student_info <- list(
    name = "Alice Smith",
    age = 21,
    is_enrolled = TRUE,
    grades = c(85, 92, 78, 95), # A numeric vector
    course_matrix = matrix(c(1, 2, 3, 4), nrow = 2, byrow = TRUE) # A matrix
)
print("Student Info List (named):")
print(student_info)

# Create a nested list (a list within a list)
nested_list_example <- list(
    item1 = "First item",
    sub_list = list(
        sub_item1 = 100,
        sub_item2 = "nested text"
    ),
    item3 = FALSE
)
print("Nested List Example:")
print(nested_list_example)

# Check the class of a list
print(class(student_info))
```

**Key Points:**

* Lists are printed with double square brackets `[[n]]` indicating the *n*th component, followed by its contents. If components are named, the name `$name` precedes the content.
* Naming list elements is highly recommended as it makes your code more readable and allows for easier access.

---

#### **Accessing List Elements**

Accessing elements in a list is a bit more nuanced than vectors or matrices due to their heterogeneous nature. R provides three main ways to access list components:

1.  **Single Square Brackets (`[]`):** Returns a *new list* containing the selected elements.
    * Useful when you want to subset the list itself, keeping its list structure.
2.  **Double Square Brackets (`[[]]`):** Returns the *content* of a single element (the actual object, not a list containing it).
    * Used to extract the actual data stored within a specific list component.
3.  **Dollar Sign (`$`):** Extracts the content of a named list element by its name.
    * Most convenient for accessing named list elements.

**Code Snippets:**

```{r}
student_data <- list(
    name = "Bob Johnson",
    id = 1056,
    scores = c(75, 80, 88),
    address = list(
        street = "123 Main St",
        city = "Anytown",
        zip = "12345"
    ),
    is_active = TRUE
)
print("Original Student Data List:")
print(student_data)

# 1. Accessing with Single Square Brackets `[]` (returns a list)
print("Accessing 'name' element as a list:")
print(student_data["name"])
print(class(student_data["name"])) # Still a list

print("Accessing 'name' and 'id' elements as a new list:")
print(student_data[c("name", "id")])

# 2. Accessing with Double Square Brackets `[[]]` (returns the element's content)
print("Accessing 'name' element content:")
print(student_data[["name"]])
print(class(student_data[["name"]])) # Returns the character vector

print("Accessing 'id' element content by position:")
print(student_data[[2]]) # id is the second element
print(class(student_data[[2]]))
```
```{r}
#### 3. Accessing with Dollar Sign `$` (returns the element's content by name)
print("Accessing 'scores' element content using $:")
print(student_data$scores)
print(class(student_data$scores))

print("Accessing 'is_active' element content using $:")
print(student_data$is_active)

# Accessing elements within a nested list
print("Accessing 'city' from nested 'address' list:")
print(student_data$address$city)
print(student_data[["address"]][["city"]]) # Alternative using double brackets

# Modifying list elements
student_data$age <- 22 # Add a new element or modify existing
student_data$scores[1] <- 78 # Modify an element within a list component
print("Student Data after modifications:")
print(student_data)

# Adding a new element to a list
student_data$major <- "Computer Science"
print("Student Data after adding 'major':")
print(student_data)

# Removing an element from a list
student_data$is_active <- NULL
print("Student Data after removing 'is_active':")
print(student_data)
```


**Key Points:**

* Use `[[]]` or `$` when you want to work *directly* with the object inside a list component (e.g., performing calculations on a numeric vector stored in a list).
* Use `[]` when you want to keep the list structure, for instance, when selecting multiple components or for subsetting.
* Lists are dynamic; you can add, modify, or remove elements after creation.

---

#### **Useful List Functions**

Several functions help you work with lists:

* `length()`: Returns the number of top-level elements in a list.
* `names()`: Returns or sets the names of list elements.
* `str()`: (Structure) Provides a compact, informative display of an R object's internal structure. Incredibly useful for lists.
* `unlist()`: Converts a list to an atomic vector. This will coerce all elements to a common data type.
* `as.list()`: Converts other R objects (like vectors) into lists.

**Code Snippets:**

```{r}
my_complex_list <- list(
    fruits = c("apple", "banana", "orange"),
    numbers = 1:5,
    matrix_data = matrix(c(1, 2, 3, 4), nrow = 2),
    is_ready = TRUE,
    nested = list(a = 1, b = "hello")
)

print("Original Complex List:")
print(my_complex_list)

# Get the number of top-level elements
print("Number of elements in list:")
print(length(my_complex_list))
```
```{r}
#### Get names of elements
print("Names of list elements:")
print(names(my_complex_list))
```
```{r}
#### Set or change names (can do this during creation or afterwards)
names(my_complex_list)[1] <- "fruits_list" # Change name of first element
print("List names after change:")
print(names(my_complex_list))
```
```{r}
#### Use str() to inspect structure (highly recommended!)
print("Structure of the list (str()):")
str(my_complex_list)

# Unlist a list (all elements will be coerced to a common type)
unlisted_data <- unlist(my_complex_list)
print("Unlisted data (all coerced to character):")
print(unlisted_data)
print(class(unlisted_data))
```
```{r}
#### Convert a vector to a list
vector_to_list <- as.list(c(10, 20, 30))
print("Vector converted to list:")
print(vector_to_list)
print(class(vector_to_list))
```

---

This lesson provided a thorough understanding of lists, their creation, and the crucial methods for accessing and manipulating their elements. Their flexibility makes them invaluable for storing complex and mixed-type data in R.

**Next, we will delve into Lesson 4: Data Frames.**
