# **Phase 1: Introduction to R and Fundamentals**

## **Module 1.2: Data Types and Structures**

---

## **Lesson 6: Data Structures Overview and Conversion**

This lesson provides a comprehensive overview of the core R data structures discussed in this module, highlighting their key characteristics. More importantly, it demonstrates how to convert between these structures, which is a common requirement in data preparation and analysis.

---

#### **Overview of R Data Structures**

Let's recap the fundamental data structures in R:

| Data Structure  | Dimensions   | Homogeneity (Data Type) | Key Characteristics                                 | Typical Use Case                                    |
| :-------------- | :----------- | :---------------------- | :-------------------------------------------------- | :-------------------------------------------------- |
| **Atomic Vector** | 1 (single row/col) | Homogeneous             | Sequence of elements of the *same* basic type.     | Storing lists of numbers, text, logical values.     |
| **Matrix** | 2 (rows & cols) | Homogeneous             | Rectangular array of elements of the *same* type.  | Linear algebra, numerical computations, small tables of uniform data. |
| **Array** | N (multi-dim) | Homogeneous             | Multi-dimensional array of elements of the *same* type. | Storing multi-dimensional data (e.g., time series across regions). |
| **List** | 1 (linear)   | Heterogeneous           | Ordered collection of objects of *any* type.       | Storing results of analyses, complex objects with varied components. |
| **Data Frame** | 2 (rows & cols) | Heterogeneous (by column) | List of equal-length vectors. Columns can be different types, rows are observations. | **The most common structure for tabular datasets.** |
| **Factor** | 1 (single row/col) | Homogeneous (numeric levels) | Represents categorical data with predefined `levels`. | Storing categorical variables (e.g., Gender, Grade). |

#### Key Principle: Homogeneity vs. Heterogeneity

* **Homogeneous:** Atomic Vector, Matrix, Array, Factor (elements must be of the same base type).
* **Heterogeneous:** List, Data Frame (can contain elements/columns of different types).

This distinction is crucial because R will automatically perform **coercion** to the most flexible type when you try to combine different types into a homogeneous structure.

---

#### **Data Structure Conversions**

Converting between data structures is a frequent task. R provides `as.type()` functions for this purpose, such as `as.vector()`, `as.matrix()`, `as.data.frame()`, `as.list()`, `as.factor()`, `as.numeric()`, `as.character()`, etc.

**Important Note on Coercion during Conversion:**
When converting from a heterogeneous structure (like a data frame or list) to a homogeneous one (like a vector or matrix), or when converting between different atomic types, R will apply its type coercion rules. This can lead to unexpected results if not understood (e.g., numbers becoming characters).

---

### **1. Vector Conversions**

* **To Numeric/Integer/Logical/Character:** `as.numeric()`, `as.integer()`, `as.logical()`, `as.character()`.

    **Code Snippets:**

    ```{r}
    # Numeric to character
    num_vec <- c(10, 20, 30)
    char_vec <- as.character(num_vec)
    print("Numeric to Character:")
    print(char_vec)
    print(class(char_vec))

    # Character to numeric (if convertible)
    num_str_vec <- c("100", "200", "300")
    converted_num_vec <- as.numeric(num_str_vec)
    print("Character (numeric strings) to Numeric:")
    print(converted_num_vec)
    print(class(converted_num_vec))

    # Character to numeric (if not convertible, results in NA and warning)
    mixed_str_vec <- c("10", "abc", "20")
    converted_mixed_vec <- as.numeric(mixed_str_vec)
    print("Character (mixed strings) to Numeric:")
    print(converted_mixed_vec)
    print(class(converted_mixed_vec))
    print(is.na(converted_mixed_vec)) # Check for NAs

    # Logical to numeric (TRUE=1, FALSE=0)
    logic_vec <- c(TRUE, FALSE, TRUE)
    num_logic_vec <- as.numeric(logic_vec)
    print("Logical to Numeric:")
    print(num_logic_vec)
    print(class(num_logic_vec))

    # Numeric (0/1) to logical
    binary_num_vec <- c(1, 0, 1, 0)
    logic_binary_vec <- as.logical(binary_num_vec)
    print("Numeric (0/1) to Logical:")
    print(logic_binary_vec)
    print(class(logic_binary_vec))
    ```

---

### **2. Matrix Conversions**

* **Vector to Matrix:** `matrix()`
* **Matrix to Vector:** `as.vector()` (or simply printing/manipulating will often coerce)
* **Matrix to Data Frame:** `as.data.frame()`

    **Code Snippets:**

    ```{r}
    # Vector to Matrix
    data_vec <- 1:6
    mat_from_vec <- matrix(data_vec, nrow = 2, byrow = TRUE)
    print("Vector to Matrix:")
    print(mat_from_vec)

    # Matrix to Vector
    vec_from_mat <- as.vector(mat_from_vec)
    print("Matrix to Vector:")
    print(vec_from_mat)
    print(class(vec_from_mat))

    # Matrix to Data Frame (columns become columns, rows become rows)
    char_matrix <- matrix(c("A", "B", "C", "D"), nrow = 2, byrow = TRUE)
    df_from_matrix <- as.data.frame(char_matrix, stringsAsFactors = FALSE)
    print("Matrix to Data Frame:")
    print(df_from_matrix)
    print(class(df_from_matrix))
    print(sapply(df_from_matrix, class)) # Check column types
    ```

---

### **3. List Conversions**

* **Vector/Matrix/Data Frame to List:** `as.list()`
* **List to Vector:** `unlist()` (beware of coercion!)
* **List to Data Frame:** `as.data.frame()` (only if all elements are vectors of equal length).

    **Code Snippets:**

    ```{r}
    # Vector to List
    vec_to_list <- as.list(c(1, 2, 3))
    print("Vector to List:")
    print(vec_to_list)
    print(class(vec_to_list))

    # List to Vector (using unlist, leads to coercion)
    my_mixed_list <- list(num = 1:3, char = c("a", "b"), logic = TRUE)
    unlisted_vec <- unlist(my_mixed_list)
    print("List to Vector (via unlist, with coercion):")
    print(unlisted_vec)
    print(class(unlisted_vec)) # All elements become character

    # List to Data Frame (if components are vectors of equal length)
    list_for_df <- list(
        Name = c("X", "Y", "Z"),
        Age = c(25, 30, 22),
        Score = c(88, 92, 75)
    )
    df_from_list <- as.data.frame(list_for_df, stringsAsFactors = FALSE)
    print("List to Data Frame:")
    print(df_from_list)
    print(class(df_from_list))

    # List to Data Frame (if components are not equal length, results in error)
    # unequal_list <- list(A = 1:3, B = 1:2)
    # as.data.frame(unequal_list) # Uncomment to see error
    ```

---

### **4. Data Frame Conversions**

* **Data Frame to Matrix:** `as.matrix()` (all columns will be coerced to a single type)
* **Data Frame to List:** `as.list()` (each column becomes a list element)
* **Data Frame to Vector:** `as.vector(unlist(your_dataframe))` (coerces all to single type)

    **Code Snippets:**

    ```{r}
    my_data_frame <- data.frame(
        ID = 1:3,
        Category = c("A", "B", "C"),
        Value = c(10.5, 20.3, 15.0),
        stringsAsFactors = FALSE
    )
    print("Original Data Frame:")
    print(my_data_frame)

    # Data Frame to Matrix (all elements coerced to character)
    matrix_from_df <- as.matrix(my_data_frame)
    print("Data Frame to Matrix (all elements become character):")
    print(matrix_from_df)
    print(class(matrix_from_df))

    # Data Frame to List (each column becomes a list element)
    list_from_df <- as.list(my_data_frame)
    print("Data Frame to List:")
    print(list_from_df)
    print(class(list_from_df))
    print(class(list_from_df$ID)) # Check type of list elements

    # Data Frame to Vector (flattened, all coerced)
    vector_from_df <- as.vector(unlist(my_data_frame))
    print("Data Frame to Vector (flattened, all coerced):")
    print(vector_from_df)
    print(class(vector_from_df)) # All elements become character
    ```

---

### **5. Factor Conversions**

* **To Character:** `as.character()` (recommended)
* **To Numeric:** `as.numeric(as.character())` (recommended for levels that are numbers)
* **To Factor:** `as.factor()`

    **Code Snippets:**

    ```{r}
    # Factor to Character
    grade_factor <- factor(c("A", "B", "C", "A", "D"), levels = c("D", "C", "B", "A"), ordered = TRUE)
    print("Original Grade Factor:")
    print(grade_factor)
    char_grade <- as.character(grade_factor)
    print("Factor to Character:")
    print(char_grade)
    print(class(char_grade))

    # Factor (with numeric levels as strings) to Numeric - IMPORTANT TWO-STEP PROCESS
    response_factor <- factor(c("1", "5", "2", "1"), levels = c("1", "2", "3", "4", "5"))
    print("Original Response Factor (numeric levels as strings):")
    print(response_factor)
    numeric_response <- as.numeric(as.character(response_factor))
    print("Factor to Numeric (correctly):")
    print(numeric_response)
    print(class(numeric_response))

    # Numeric/Character to Factor
    region_data <- c("East", "West", "North", "East")
    region_factor <- as.factor(region_data)
    print("Character to Factor:")
    print(region_factor)
    print(class(region_factor))
    print(levels(region_factor))
    ```


---

This lesson summarized the key R data structures and, critically, explored how to convert between them. Understanding the nuances of coercion during conversion is vital to avoid data loss or unexpected results. You now have a solid foundation in R's fundamental data structures.
